A Practical Optional Type System for Clojure

### 2012-12-06 dissertation/intro.tex
Fix ref sorting. Adjust listings to not break pages

By adding static type checking to Clojure, we can achieve errors messages that are closer to
Haskell's.
Listing \ref{lst:tcctime} gives an idea of what kind of error messages Typed Clojure actually
provides in these situations.

@@

Listing \ref{lst:tcctime} deserves some explanation:

\begin{itemize}
  \item The given source of the error may be surprising, but it is explained
        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
        \lstinline|(. clojure.lang.Numbers add "string" a)|.
        \lstinline|cf| type checks a form.
  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
  \item \lstinline|cf| type checks a form.
  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
\end{itemize}

@@

The proliferation of other abstract programming styles, like monads, in statically typed languages
also suggested that such a static type system could
help the programmer write and debug such code, while improving confidence that their code is correct.

\subsection{Why does not Clojure feature static type checking?}

Given the advantages of static type checking outlined in the previous section, it is natural to ask:
why does Clojure not already support static type checking?
The simplest answer is that Clojure was not designed to support a static type system.
Instead, features and idioms that are not necessarily compatible with traditional type systems
were favoured, like variable-arity functions, untyped hash-maps used like objects, and extension of program behaviour
at runtime.

Without specifically speaking for the designers of Clojure, at the time Clojure was designed
it was not clear how to reconcile Clojure's feature-set with static type checking in a satisfying way.
Since then, Typed Racket, which add static type checking to the Racket programming language,
have matured and gained acceptance.
Once this occurred, it seemed like only a matter of time until the ideas developed for the Typed Racket project 
were tested for their applicability to Clojure. 
As far as we know, this project is the first serious attempt to build an optional static type system for Clojure.

@@

  \item We support safer interoperability with Java. We show how techniques developed for
        Typed Racket, such as occurrence typing~\cite{TF10},  of Java's \emph{null}
        while still conforming to existing Clojure idioms.

@@

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces~\cite{InterfaceTutorial} and datatypes are similar to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.

\begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
(ns typed.test.deftype
  (:require [typed.core 
             :refer [check-ns ann-datatype
                     tc-ignore ann-protocol AnyInteger]]))

(ann-protocol Age 
  :methods
  {age [Age -> AnyInteger]})
(tc-ignore
  (defprotocol Age
    (age [this])))

(ann-datatype Person 
  [[name :- String]
   [age :- AnyInteger]])
(deftype Person [name age]
  Age
  (age [this] age))

@@

\lstinline|defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
with a set of methods. \lstinline|ann-protocol| is a Typed Clojure macro for annotating 
a protocol with the types of its methods.
In this example, we define a protocol \lstinline|Age| with an \lstinline|age| method,
which is really a first-class function taking the target object as the first parameter. The
type signature provided with \lstinline|ann-protocol|, here \lstinline|[Age -> AnyInteger]|
which is the type of the function taking an \lstinline|Age| and returning an \lstinline|AnyInteger|,
is for this function.

Invocations of \lstinline|defprotocol| are currently not able to be type checked
and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.

@@

Typed Clojure provides heterogeneous map types which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.

\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
...
(ann config '{:file String, :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

@@

\subsubsection{Uniform Variable-Arity}

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al.\ attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.

\begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
(ann not= (Fn [Any -> boolean]
              [Any Any -> boolean]
              [Any Any Any * -> boolean]))
(defn not=
  "Same as (not (= obj1 obj2))"
  ([x] false)
  ([x y] (not (= x y)))
  ([x y & more]
   (not (apply = x y more))))
\end{lstlisting}

@@

For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences in a pair-wise fashion.
Its type is given in listing \ref{lst:maptype}.

@@

\begin{lstlisting}
[[boolean String Integer -> Number] (U (Seqable boolean) nil) (U (Seqable String) nil) (U (Seqable Integer) nil) -> (LazySeq Number)]
\end{lstlisting}

@@

Typed Clojure supports integration with Java's type system during
interoperability from Clojure. This offers many of the same advantages
as using Java's type system from Java, for example it is a type
error to pass arguments of the wrong type to methods.
Typed Clojure also makes different decisions to Java's type system,
particularly in the treatment of Java's \emph{null}.

From the perspective of static types in Java, \emph{null} is included in all reference types.
\emph{null} is represented in Clojure by the value \lstinline|nil|. Unlike Java's type system
Typed Clojure explicitly separates \emph{null} and reference types giving Typed Clojure 
more accurate types: we can directly express \emph{nullable} and \emph{not nullable} types.
(A type is \emph{nullable} if it may also be \emph{null},
which is expressed in Typed Clojure by creating a union of the type and \lstinline|nil|.
A type is \emph{not nullable} if it doesn't include \emph{null}).

\begin{lstlisting}[caption=Java interoperability with Typed Clojure]
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

@@

This example shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}.

>>>>

By adding static type checking to Clojure, we can achieve errors messages that are closer to
Haskell's.
Listing \ref{lst:tcctime} gives an idea of what kind of error messages Typed Clojure actually
provides in these situations. 
However, it deserves some explanation:

\begin{itemize}
  \item The given source of the error may be surprising, but it is explained
        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
        \lstinline|(. clojure.lang.Numbers add "string" a)|.
  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
  \item \lstinline|cf| type checks a form.
  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
\end{itemize}

@@

Most importantly: the error was caught at \emph{compile-time}.

@@

Programmers in statically typed languages take advantage of the static type system
to help verify code written in abstract programming styles, like monads.
Adding a static type system to Clojure should bring similar benefits to Clojure programmers.

\subsection{Why does Clojure not already feature static type checking?}

Given the advantages of static type checking outlined in the previous section, it is natural to ask:
why does Clojure not already support static type checking?
The simplest answer is that Clojure was not designed to support a static type system.
Instead, features and idioms that are not necessarily compatible with traditional type systems
were favoured, like variable-arity functions, untyped hash-maps used like objects, and dynamic extension of program behaviour.

Without specifically speaking for the designers of Clojure, at the time Clojure was designed
it was not clear how to reconcile Clojure's feature-set with static type checking in a satisfying way.
Since then, Typed Racket, which add static type checking to the Racket programming language,
has matured and gained acceptance.
Once this occurred, it seemed only a matter of time until the ideas developed for the Typed Racket project 
were tested for their applicability to Clojure. 
As far as we know, this project is the first serious attempt to build an optional static type system for Clojure.

@@

  \item We support safer interoperability with Java from Clojure. We show how techniques developed for
        Typed Racket, such as occurrence typing~\cite{TF10}, help detect misuses of Java's \emph{null}
        while still conforming to existing Clojure idioms.

@@

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces~\cite{InterfaceTutorial} and datatypes are similar to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.
%bump over the page
\\\\\\\\\\
\begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
(ns typed.test.deftype
  (:require 
    [typed.core :refer [check-ns ann-datatype defprotocol>
                        ann-protocol AnyInteger]]))

(ann-protocol Age 
  age [Age -> AnyInteger])
(defprotocol> Age
  (age [this]))

(ann-datatype Person 
  [name :- String]
   age :- AnyInteger])
(deftype Person [name age]
  Age
  (age [this] age))

@@

\lstinline|defprotocol>| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
with a set of methods. It is identical to the usual \lstinline|defprotocol|, but is required when using Typed Clojure
due to its complicated macroexpansion.
\lstinline|ann-protocol| is a Typed Clojure macro for annotating 
a protocol with the types of its methods.
In this example, we define a protocol \lstinline|Age| with an \lstinline|age| method,
which is really a first-class function taking the target object as the first parameter. The
type signature provided with \lstinline|ann-protocol|, here \lstinline|[Age -> AnyInteger]|
which is the type of the function taking an \lstinline|Age| and returning an \lstinline|AnyInteger|,
is for this function.

@@

Typed Clojure provides heterogeneous map types which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.
%bump
\\\\
\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
(ann config '{:file String, :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
\end{lstlisting}

@@

\subsubsection{Uniform Variable-Arity Functions}

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al.\ attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.
\\\\\\\\

%bump over page
\begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
(ann not= (Fn [Any -> boolean]
              [Any Any -> boolean]
              [Any Any Any * -> boolean]))
(defn not=
  "Same as (not (= obj1 obj2))"
  ([x] false)
  ([x y] (not (= x y)))
  ([x y & more]
   (not (apply = x y more))))
\end{lstlisting}

@@

For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
and returns a sequence of the results of applying the function to each element of the sequences simultaneously.
Its type is given in listing \ref{lst:maptype}.

@@

\begin{lstlisting}
[[boolean String -> Number] (U (Seqable boolean) nil) 
  (U (Seqable String) nil) -> (LazySeq Number)]
\end{lstlisting}

@@

Typed Clojure supports integration with Java's type system during
interoperability from Clojure. This offers many of the same advantages
as using Java's type system from Java, for example it is a type
error to pass arguments of the wrong type to methods.
Typed Clojure also makes different decisions to Java's type system,
particularly in the treatment of Java's \emph{null}.
Note that these decisions do not change Java's type system, but only
how Typed Clojure handles Java interoperability compared to how someone
familiar with Java might expect.

From the perspective of static types in Java, \emph{null} is included in all reference types.
\emph{null} is represented in Clojure by the value \lstinline|nil|. Unlike Java's type system
Typed Clojure explicitly separates \emph{null} and reference types giving Typed Clojure 
more accurate types: we can directly express \emph{nullable} and \emph{not nullable} types.
(A type is \emph{nullable} if it may also be \emph{null},
which is expressed in Typed Clojure by creating a union of the type and \lstinline|nil|.
A type is \emph{not nullable} if it doesn't include \emph{null}).
%bump over the page
\\\\
\begin{lstlisting}[caption=Java interoperability with Typed Clojure, label=lst:interopfile]
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

@@

Listing \ref{lst:interopfile} shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}.
